### 2023-02-13 article.tex
Do spell check

% Alex and Blake?
Two different researchers in genetic epideomology (more on that field later), 
write two equally good manuscripts
that describe an experiment with computational steps.
Both manuscripts are accepted by an equally prestigious journal after peer review. 
One researcher, however, does not supply the
computer code (from now on: 'code') that was used to generate the results,
where the other does.
Are the conclusion of these papers to be trusted equally?
Is this difference relevant and worth the effort?
How common is it to share code, and, if shared, how can it be preserved?
This chapter discusses why code, a type of paradata, 
should be supplied and what features it needs to have
for it to be useful.

@@

A recent paper explores the use of paradata to increase the impact of data,
stating that lack of paradata can be seen as 'a drastic constraint'
in the use of data and offer some suggestions to 
make paradata useful for data re(use):
paradata should be comprehensive, documented in a useful way, 
the documention and data should have co-evolved, 
and the paradata should be computer-friendly \cite{huvila2022improving}.
At first glance, these suggestions appear to work well for code
and will be discussed in detail below.

@@

To determine how useful code is, it needs to be available.
However, it is not common to publish the code of an experiment or analysis 
\cite{stodden2011trust,read2015sizing} (with a pleasant exception 
being \cite{conesa2019making}).
For example, in computer graphics, 
a field intimately familiar with computer code,
42\% of 454 SIGGRAPH papers supply computer code \cite{bonneel2020code}.
Another study analysed the reproducibility of registered reports
in the field of psychology, 
where 60\% of 62 studies supplied the code 
to redo the analysis \cite{obels2020analysis}.
For articles in Science magazine, 12\% of 204
studies published the code \cite{stodden2018empirical} 
(note that these were years 2011-2012).
Unpublished code has been the cause of some saddening examples,
such as an algorithm that detects breast cancer from images 
better than a human expert, 
yet failed to ever be reproduced \cite{haibe2020importance}.

@@

When code of an academic paper is not published, 
one could contact the correspending author and request it.
However, the reponse rate of correnspondence authors 
with a request for data in computational fields is around 50\% 
\cite{manca2018non, stodden2018empirical, teunis2015corresponding} 
(the field of emergency medicine seems to be a pleasant outlier, 
with 73\% of 118 emails being replied \cite{o2003email}).
When getting an answer of the corresponding author, 
48\% out of 134 replied emails will actually result 
in a sharing of the code \cite{stodden2018empirical}.
The responses of unwilling authors (see \cite{stodden2018empirical} for 
some real examples) can come across as so caustic, 
that one may be excused from not contacting an a corresponding author.

@@

\iffalse
\begin{figure}[!htbp]
  \centering
  \includegraphics[width=\linewidth]{1189px-Eukaryote_DNA-en.png}
  \caption{
    A cell has a nucleus that contains chromosomes. 
    Each of these chromosomes (46 in humans) consist out of DNA. 
    DNA itself conists out of 4 nucleotides, 
    as depicted by the horizontal sticks 
    with the colors red, yellow, green and blue.
    From \cite{sponk2012}
  }
  \label{fig:eukakyote_dna}
\end{figure}
\fi

The first type of primary data, the genotypes, 
consists out of single nucleotide polymorphisms (SNPs, pronounced 'snips').
A SNP has a name and a location within the genome.
At the SNP's location in the DNA, there will be two nucleotides.
One of these nucleotides is inherited from the mother, 
the other from the father.
The DNA consists out of billions of nucleotides.
There are four types of nucleotides,
called adenosine, cytosine, guanine and thyrosine, 
commonly abreviated as A, C, G and T 
\iffalse
(see figure \ref{fig:eukakyote_dna}))
\fi
respectively.

@@

\iffalse
\begin{figure}[!htbp]
  \centering
  \includegraphics[width=\linewidth]{ahsan2017relative_table_2_sub.png}
  \caption{
    An example result of a genetic epideological research.
    It shows that the SNP named rs12133641 (located at position 154,428,283
    of chromosome 1) is highly correlated (p value is 3.0 * 10e-73, 
    961 individuals) to the concentration of the protein IL6RA, as measured
    in blood. The table is a simplified result from \cite{ahsan2017relative}.
  }
  \label{fig:ahsan2017relative_table_2_sub}
\end{figure}
\fi

@@

From this simple example, it will be easier
to measure the level of IL6RA in the blood, than using a gene panel,
as blood tests are easier and cheaper.
However, there are associations published for many diseases,
in which one SNP (e.g. phenylketonuria) or many SNPs (e.g. \cite{bruce2009metabolic}) 
contribute to being more likely to develope a disease in the future. 
Here, the phenotype (having a disease in the future) 
is impossible to detect at the present
and associations found in earlier studies are used to create a gene panel.
As creating a gene panel is costly, those associations better be correct.

@@

Note that there is another side of this same coin:
correctly reporting the t-test in a reproducable ways
is known to be a problem, with only 80 out of a selected 179 papers
fully reporting the parity of t-test (i.e. paired or unpaired) 
\cite{weissgerber2018we}.
In those unresolved 99 papers, 
the code (if published) would clarify this.

@@

Some code hosts, among others, GitHub, 
allow the researcher/user to start software scripts upon uploading a change.
This is useful to determine the reproducibility of the code: 
one such script should run the code (on trivial data) 
to verify if the code works at all.
If the functionality of the code is lost, the researcher
can immediatly see which small change was the cause of this.
Using a CI service (such as GitHub Actions for GitHub), 
is known to significantly 
increase the number of bugs exposed \cite{vasilescu2015} and increases
the speed at which new features are added \cite{vasilescu2015}.
Using CI opens up the possibility to informally annotate features of
a repository. 
\iffalse
For example, Figure \ref{fig:build_badge} shows a 
build badge that indicates that a \LaTeX document (i.e. this paper)
could be built.
\fi

@@

The percentage of (lines of) code tested is called the code coverage.
Code coverage correlates with code quality \cite{horgan1994,del1995correlation}. 
The non-profit organisation rOpenSci \cite{ram2018community},
which peer-reviews code,
has made it a prerequisite to have a code coverage of 100\%.
CI opens up the possibility to informally annotate the
code coverage. 
For example, build badges
\iffalse
(e.g. as shown in Figure \ref{fig:badge_codecov})
\fi
are visual indications on a website that can, among others,
show a certain R package's code has a 100\% code coverage.

@@

In a perfect world, all code has the characteristis of ideal paradata
and is written from software development best practices.
This section discusses the problems that arise by doing so.

@@

Code needs maintanance:
the libraries that are used by an experiment
will change over time, inevitably causing the experiment to fail
over time. 
Code that will stand the test of time perfectly 
is deemed 'impossible' \cite{benureau2018re}.

@@

However, upon the find of a new bug, the question is:
does all the research that use that software still 
result in the same conclusions?
To prevent authors chosing to hide bug reports, 
a code hosting website usually has a public place to report bugs,
to increase transparency in the handling of bug reports.
Should there be a retraction notice on code?

@@

Publishing code may be disadvantageous for an author.
For science, code should be published, 
this allows reproducible research 
(see \cite{haibe2020importance} for a tragic example).
For an author, publishing code alongside an experiment opens up
the possibily to receive questions regarding that code.
Note, however, that not publishing code may put 
oneself in the focus of attention
and -after much effect by others reproducing an incorrect result-
at the cost of a scientific career \cite{baggerly2009deriving}.

@@

When applying these general recomendations to code, 
this list can be phrased more precisely:

@@

\begin{table}[h]
  \begin{tabular}{|p{2cm}|l|}
    \hline
    \textbf{Type of code} & \textbf{Recommendation} \\
    \hline
    No code       & Disallow papers that do not supply code publicly \\
    \hline
    Raw code      & Don't publish raw code as text, host it on a code hosting website \\
    \hline
    Hosted code   & Use a standarized badge for build status \\
                  & Use a standarized badge for code coverage \\
                  & Standardize to link between paper and code host \\
    \hline
    Runnable code & Allow to upload containers without limits \\
                  & Website to host containers must allow for annotation \\
                  & Use a standarized badge for the container to be functioning  \\
                  & Standarize to link between container, paper and code repository \\
    \hline
  \end{tabular}
  \caption{Recommendations described in this paper}
  \label{tab:recommendations}
\end{table}

>>>>

% Alex and Blake?
Two different researchers in genetic epidemiology (more on that field later), 
write two equally good manuscripts
that describe an experiment with computational steps.
Both manuscripts are accepted by an equally prestigious journal after peer review. 
One researcher, however, does not supply the
computer code (from now on: 'code') that was used to generate the results,
where the other does.
Are the conclusion of these papers to be trusted equally?
Is this difference relevant and worth the effort?
How common is it to share code, and, if shared, how can it be preserved?
This chapter discusses why code, a type of paradata, 
should be supplied and what features it needs to have
for it to be useful.

@@

A recent paper explores the use of paradata to increase the impact of data,
stating that lack of paradata can be seen as 'a drastic constraint'
in the use of data and offer some suggestions to 
make paradata useful for data re(use):
paradata should be comprehensive, documented in a useful way, 
the documentation and data should have co-evolved, 
and the paradata should be computer-friendly \cite{huvila2022improving}.
At first glance, these suggestions appear to work well for code
and will be discussed in detail below.

@@

To determine how useful code is, it needs to be available.
However, it is not common to publish the code of an experiment or analysis 
\cite{stodden2011trust,read2015sizing} (with a pleasant exception 
being \cite{conesa2019making}).
For example, in computer graphics, 
a field intimately familiar with computer code,
42\% of 454 SIGGRAPH papers supply computer code \cite{bonneel2020code}.
Another study analyzed the reproducibility of registered reports
in the field of psychology, 
where 60\% of 62 studies supplied the code 
to redo the analysis \cite{obels2020analysis}.
For articles in Science magazine, 12\% of 204
studies published the code \cite{stodden2018empirical} 
(note that these were years 2011-2012).
Unpublished code has been the cause of some saddening examples,
such as an algorithm that detects breast cancer from images 
better than a human expert, 
yet failed to ever be reproduced \cite{haibe2020importance}.

@@

When code of an academic paper is not published, 
one could contact the corresponding author and request it.
However, the response rate of corresponding authors 
with a request for data in computational fields is around 50\% 
\cite{manca2018non, stodden2018empirical, teunis2015corresponding} 
(the field of emergency medicine seems to be a pleasant outlier, 
with 73\% of 118 emails being replied \cite{o2003email}).
When getting an answer of the corresponding author, 
48\% out of 134 replied emails will actually result 
in a sharing of the code \cite{stodden2018empirical}.
The responses of unwilling authors (see \cite{stodden2018empirical} for 
some real examples) can come across as so caustic, 
that one may be excused from not contacting an a corresponding author.

@@

\iffalse
\begin{figure}[!htbp]
  \centering
  \includegraphics[width=\linewidth]{1189px-Eukaryote_DNA-en.png}
  \caption{
    A cell has a nucleus that contains chromosomes. 
    Each of these chromosomes (46 in humans) consist out of DNA. 
    DNA itself consists out of 4 nucleotides, 
    as depicted by the horizontal sticks 
    with the colors red, yellow, green and blue.
    From \cite{sponk2012}
  }
  \label{fig:eukakyote_dna}
\end{figure}
\fi

The first type of primary data, the genotypes, 
consists out of single nucleotide polymorphisms (SNPs, pronounced 'snips').
A SNP has a name and a location within the genome.
At the SNP's location in the DNA, there will be two nucleotides.
One of these nucleotides is inherited from the mother, 
the other from the father.
The DNA consists out of billions of nucleotides.
There are four types of nucleotides,
called adenosine, cytosine, guanine and thyrosine, 
commonly abbreviated as A, C, G and T 
\iffalse
(see figure \ref{fig:eukakyote_dna}))
\fi
respectively.

@@

\iffalse
\begin{figure}[!htbp]
  \centering
  \includegraphics[width=\linewidth]{ahsan2017relative_table_2_sub.png}
  \caption{
    An example result of a genetic epidemiological research.
    It shows that the SNP named rs12133641 (located at position 154,428,283
    of chromosome 1) is highly correlated (p value is 3.0 * 10e-73, 
    961 individuals) to the concentration of the protein IL6RA, as measured
    in blood. The table is a simplified result from \cite{ahsan2017relative}.
  }
  \label{fig:ahsan2017relative_table_2_sub}
\end{figure}
\fi

@@

From this simple example, it will be easier
to measure the level of IL6RA in the blood, than using a gene panel,
as blood tests are easier and cheaper.
However, there are associations published for many diseases,
in which one SNP (e.g. phenylketonuria) or many SNPs (e.g. \cite{bruce2009metabolic}) 
contribute to being more likely to develop a disease in the future. 
Here, the phenotype (having a disease in the future) 
is impossible to detect at the present
and associations found in earlier studies are used to create a gene panel.
As creating a gene panel is costly, those associations better be correct.

@@

Note that there is another side of this same coin:
correctly reporting the t-test in a reproducible ways
is known to be a problem, with only 80 out of a selected 179 papers
fully reporting the parity of t-test (i.e. paired or unpaired) 
\cite{weissgerber2018we}.
In those unresolved 99 papers, 
the code (if published) would clarify this.

@@

Some code hosts, among others, GitHub, 
allow the researcher/user to start software scripts upon uploading a change.
This is useful to determine the reproducibility of the code: 
one such script should run the code (on trivial data) 
to verify if the code works at all.
If the functionality of the code is lost, the researcher
can immediately see which small change was the cause of this.
Using a CI service (such as GitHub Actions for GitHub), 
is known to significantly 
increase the number of bugs exposed \cite{vasilescu2015} and increases
the speed at which new features are added \cite{vasilescu2015}.
Using CI opens up the possibility to informally annotate features of
a repository. 
\iffalse
For example, Figure \ref{fig:build_badge} shows a 
build badge that indicates that a \LaTeX document (i.e. this paper)
could be built.
\fi

@@

The percentage of (lines of) code tested is called the code coverage.
Code coverage correlates with code quality \cite{horgan1994,del1995correlation}. 
The non-profit organization rOpenSci \cite{ram2018community},
which peer-reviews code,
has made it a prerequisite to have a code coverage of 100\%.
CI opens up the possibility to informally annotate the
code coverage. 
For example, build badges
\iffalse
(e.g. as shown in Figure \ref{fig:badge_codecov})
\fi
are visual indications on a website that can, among others,
show a certain R package's code has a 100\% code coverage.

@@

In a perfect world, all code has the characteristics of ideal paradata
and is written from software development best practices.
This section discusses the problems that arise by doing so.

@@

Code needs maintenance:
the libraries that are used by an experiment
will change over time, inevitably causing the experiment to fail
over time. 
Code that will stand the test of time perfectly 
is deemed 'impossible' \cite{benureau2018re}.

@@

However, upon the find of a new bug, the question is:
does all the research that use that software still 
result in the same conclusions?
To prevent authors choosing to hide bug reports, 
a code hosting website usually has a public place to report bugs,
to increase transparency in the handling of bug reports.
Should there be a retraction notice on code?

@@

Publishing code may be disadvantageous for an author.
For science, code should be published, 
this allows reproducible research 
(see \cite{haibe2020importance} for a tragic example).
For an author, publishing code alongside an experiment opens up
the possibility to receive questions regarding that code.
Note, however, that not publishing code may put 
oneself in the focus of attention
and -after much effect by others reproducing an incorrect result-
at the cost of a scientific career \cite{baggerly2009deriving}.

@@

When applying these general recommendations to code, 
this list can be phrased more precisely:

@@

\begin{table}[h]
  \begin{tabular}{|p{2cm}|l|}
    \hline
    \textbf{Type of code} & \textbf{Recommendation} \\
    \hline
    No code       & Disallow papers that do not supply code publicly \\
    \hline
    Raw code      & Don't publish raw code as text, host it on a code hosting website \\
    \hline
    Hosted code   & Use a standardized badge for build status \\
                  & Use a standardized badge for code coverage \\
                  & Standardize to link between paper and code host \\
    \hline
    Runnable code & Allow to upload containers without limits \\
                  & Website to host containers must allow for annotation \\
                  & Use a standardized badge for the container to be functioning  \\
                  & Standardize to link between container, paper and code repository \\
    \hline
  \end{tabular}
  \caption{Recommendations described in this paper}
  \label{tab:recommendations}
\end{table}
